---
title: "SurvivalAnalysis: From life table to survival curves"
description: |
  Hands on Survival Analysis
author: Ming
date: 2022-01-02
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F,warning = F) 
```


# Data organization

## Dlife data structure explained

Dlife is a software for lifespan experiment.

* Censensus is performed every other day.
* setup: vial ID, #total.fly per vial
* every censunsus, record: #time, #deaths, #carried, #censored

The example dataset below contain lifespan data from 16 vials of 2 genotypes.

```{r}
library(ggplot2);library(data.table)
x<-readxl::read_xlsx("Tau_lifespan_Aug_19_2021.xlsx",sheet='RawData') #DL=DLife, this data contain 48 vial lifespan data
table(x$Chamber,x$UniqueName)
x=as.data.table(x)

DL=x[x$Chamber==1,] #select vial 1 as an example
head(DL)
DL[AgeH==1,]$N #Total number of flies in the experiment 

# select informative rows where events happen
InfoDL<-DL[(Deaths>0|Censored>0 | IntDeaths>0), .(AgeH, Chamber, UniqueName, Deaths, Carried, Censored, IntDeaths, N)]

# assmue t1: previous time
#        t2: current time
# IntDeath.t2 = Deaths.t2 + Carried.t2 - Carried.t1
# N.t2 = N.t1 - Censored.t2 - IntDeath.t2
plot(InfoDL$AgeH,InfoDL$N,type='l',xlab='Time in hr',ylab='#fly')
```

# Kaplan Meier Analysis

Useful link: http://www.sthda.com/english/wiki/survival-analysis-basics

## Reformat data for survival R package input

```{r}
library(survival)

## Convert DLife data to the format usable by the survival package. 
# survival package require each death correspond to one row
# input to survival package:
# event happening time point, which event (death or census)
DL<-readxl::read_xlsx("Tau_lifespan_Aug_19_2021.xlsx",sheet='RawData') #DL=DLife, this data contain 48 vial lifespan data
DL=as.data.table(DL)
min(DL$AgeH)
DL[AgeH==min(DL$AgeH), sum(N)] #Total number of flies in the experiment 

# select informative rows where events happen
InfoDL<-DL[(Censored>0 | IntDeaths>0), .(AgeH, Chamber, UniqueName, Deaths, Carried, Censored, IntDeaths, N)]
# select non-redundant columns used in survival analysis
eDL<-InfoDL[, .(AgeH, Chamber, UniqueName, Censored, IntDeaths)]
dim(eDL) #181 rows

library(splitstackshape) 
sum(eDL$IntDeaths) #360
Deaths<-expandRows(eDL[,.(AgeH, Chamber, UniqueName, IntDeaths)], "IntDeaths")
dim(Deaths) #360
Deaths$Deaths<-1 #Assign value 1 to each expanded row
Deaths$Censored<-0 #Assign value 0 to each expanded row

sum(eDL$Censored) #24
Censored<-expandRows(eDL[,.(AgeH, Chamber, UniqueName, Censored)], "Censored")
dim(Censored) #24
Censored$Censored<-1
Censored$Deaths<-0

temp<-rbind(Deaths, Censored) #temp$Deaths is equivalent to event in the survival object, as all deaths are 1 and all censored are 0
Data<-temp[,Genotype:=UniqueName]
Data$AgeD<-(Data$AgeH)/24 #hours into days

#write.table(Data, "Surv_input.txt", sep="\t", quote=F, row.names=F)

## survfit
fit<-survfit(Surv(AgeD,Deaths)~Genotype, data=Data)
summary(fit)$table

##Survival estimates at an early age
head(fit$time)
SurvTo40<-summary(fit, times=40)$surv
SurvTo40

ndeath<-summary(fit)$table[,4]
rmean<-summary(fit)$table[,5]
se<-summary(fit)$table[,6]
median<-summary(fit)$table[,7]
SummaryStats<-data.table(Genotype=levels(as.factor(Data$Genotype)),NDeath=ndeath,Mean=rmean,Se=se,Median=median)
#write.table(SummaryStats, "Surv_LifeSpan_Summary", row.names=F, sep="\t", quote=F)


# visualizae fitting curves https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_surv.html
library(ggfortify)
autoplot(fit)


# log-rank test
#https://zhuanlan.zhihu.com/p/138280210
surv_diff <- survdiff(Surv(AgeD, Deaths) ~ Genotype, data = Data)
surv_diff
```


# Parametric Survival Analysis

## Exponential model
```{r}
res.exp <- survreg(Surv(AgeD, Deaths) ~ Genotype, data = Data, dist = "exponential")
summary(res.exp)
```

## Weibull model
```{r}
res.weibull <- survreg(Surv(AgeD, Deaths) ~ Genotype, data = Data, dist = c("weibull", "exponential", "gaussian", "logistic", 
    "lognormal", "loglogistic")[1])
summary(res.weibull)
```

## Cox proportional hazards model
```{r}
# genotype as fixed effect
fit.fixed<-coxph(Surv(AgeD,Deaths)~Genotype, data=Data)
summary(fit.fixed)

# genotype as random effect
library(coxme)
fit.random1<-coxme(Surv(AgeD,Deaths)~(1|Genotype), data=Data)
fit.random1

fit.random2<-coxme(Surv(AgeD,Deaths)~(1|Chamber), data=Data)
fit.random2


fit.nested<-coxme(Surv(AgeD,Deaths)~Genotype+(1|Chamber), data=Data)
fit.nested


res.cox <- coxph(Surv(AgeD, Deaths) ~ Genotype, data = Data)
summary(res.cox)
```

## Gompertz model

```{r}
# use library(flexsurv) for fitting gompertz and estimating parameters
library(flexsurv)
# one sample
unique(Data$Genotype)
subData<-Data[Genotype=="w1118"]
fs1<-flexsurvreg(Surv(AgeD,Deaths)~1, data=subData, dist="gompertz")
fs1$coefficients

# prepare a matrix for storing fitting Gompertz results
Gompertz<-matrix(NA, nrow=length(levels(as.factor(Data$Genotype))), ncol=2)
rownames(Gompertz)<-levels(as.factor(Data$Genotype))
colnames(Gompertz)<-c("shape","rate")

for (i in 1:length(levels(as.factor(Data$Genotype)))) {
  subData<-Data[Genotype==levels(as.factor(Data$Genotype))[i]]
  fs<-flexsurvreg(Surv(AgeD,Deaths)~1, data=subData, dist="gompertz")
  Gompertz[i,]<-fs$coefficients ##Extract alpha ("rate") and beta ("shape") from the model fit
}

Gompertz #every genotype has a shape and a rate estimated values


res.gompertz <- flexsurvreg(Surv(AgeD, Deaths) ~ Genotype, data = Data, dist = c("gompertz"))
res.gompertz
```


## plot fitting curves 

```{r}
#http://www.stat.columbia.edu/~madigan/W2025/notes/survival.pdf
## Define a function to plot survreg prediction by gender
survreg.curves <- function(model, col = "black", values = c("ON4R","w1118"),seq.quantiles = seq(0.00, 1.00, 0.01)) {
  
  plyr::l_ply(values, function(X) {
    if(X=='tau'){type=1}else{type=2}
    lines(x = predict(model,                    # survreg object to use
                      newdata=data.frame(Genotype = X), # Dataset to perform prediction for
                      type = "quantile",                # Predict survival time (X-axis values) given event quantile
                      p = seq.quantiles),               # Vector of quantiles (Y-axis values)
          y = (1 - seq.quantiles),              # Change to survival quantile (proportion remaining)
          col = col, lty = type)               # COLor and Line TYpe
  })
}

## plot KM curves
par(lwd = 2)
plot(fit, lty = 1:2, conf.int = F, , main = "nsyb, male", xlab = "Days", ylab = "Survival")
legend("topright", c("ON4R", "w1118"), lty = 1:2)
## plot Cox prediction
lines(survfit(res.cox, newdata = data.frame(Genotype = c("ON4R", "w1118"))), col = "green", lty = 1:2, mark.time = FALSE)
## Plot exponential model prediction
survreg.curves(res.exp, "purple")
## Plot Weibull model prediction
survreg.curves(res.weibull, "blue")

## Plot Gompertz model prediction
out = summary(res.gompertz)
lines(out$`Genotype=w1118`[, 1], out$`Genotype=w1118`[, 2], col = "red", lty = 2)
lines(out$`Genotype=ON4R`[, 1], out$`Genotype=ON4R`[, 2], col = "red", lty = 1)

res.gompertz$res

## Add legends
legend(x = "bottomleft", legend = c("Kaplan-Meier", "Cox", "Exponential", "Weibull", "Gompertz"), lwd = 2, bty = "n", col = c("black", 
    "green", "purple", "blue", "red"))

```


### More abnout Gompertz distribution  

<br>
* The Gompertz distribution with shape parameter `a` and rate parameter `b` has 
    + PDF, probability density function: f(x|a,b) = b exp(ax) exp(-b/a (exp(ax) - 1))

    + hazard function: h(x|a,b) = b exp(ax)
    + CDF: F(x|a,b) = 1 - exp( -b/a (exp(at)-1))
    + mortality function m(t), log of h(t): m(t) = log(b) + a*x

b: rate
<br>
a: shape

#### Fit Tau and Control flies separately and get the coeff estimates


```{r results="markup",echo=T,fig.height=6, fig.width=8}
## Gompertz model
control=subset(Data,Genotype=='w1118')
res.gompertz1 <- flexsurvreg(Surv(AgeD,Deaths)~1,data=control,
                                dist= c("gompertz"))
res.gompertz1

tau=subset(Data,Genotype=='ON4R')
res.gompertz2 <- flexsurvreg(Surv(AgeD,Deaths)~1,data=tau,
                                dist= c("gompertz"))
res.gompertz2

a=res.gompertz1$res[1];
b=res.gompertz1$res[2];

t=seq(1,100,1);
y=1-exp(-b/a*(exp(a*t)-1)) #F(t)
plot(t,1-y,xlab="time",ylab="survival",col="blue",ylim=c(0,1))

a=res.gompertz2$res[1]; #shape <=> slope for motality func
b=res.gompertz2$res[2]; #rate <=> log(rate)=intercept for motality func

t=seq(1,100,1);
y=1-exp(-b/a*(exp(a*t)-1)) #S(t)
points(t,1-y,col="red")
legend(x = "bottomleft",
           legend = c("w1118","Tau"),
           lwd = 2, bty = "n",
           col = c("blue","red"))
plot(x=-20:20,y=-20:20,type="n",ylab="mortality (log of h(t))")
abline(a=res.gompertz1$res.t[2,1],b=res.gompertz1$res.t[1,1],col="blue") #a=intercept;b=slope
abline(a=res.gompertz2$res.t[2,1],b=res.gompertz2$res.t[1,1],col="red")

plot(x=0:100,y=-50:50,type="n",ylab="mortality (log of h(t))")
abline(a=res.gompertz1$res.t[2,1],b=res.gompertz1$res.t[1,1],col="blue") #a=intercept;b=slope
abline(a=res.gompertz2$res.t[2,1],b=res.gompertz2$res.t[1,1],col="red")

x=data.frame(rbind(res.gompertz1$res.t[,1],res.gompertz2$res.t[,1]));
colnames(x)=c("slope",'intercept');
rownames(x)=c("w1118","tau");
x

```
```{r}
devtools::session_info()
```
