---
title: "TechNote: Enable OpenMP on macOS Mojave"
description: |
  Parallel Computing in R
author: Ming
date: 2021-12-14
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F,warning = F) 
```


# Setup: openMP realization on mac 

When I was running `sourceCpp` in Rstudio in my mac laptop, it showed that [OpenMP](https://mac.r-project.org/openmp/) was not detected in R .

For example, a piece of Rcpp code from [hogsvd.cpp](https://github.com/barkasn/hogsvdR/blob/master/src/hogsvd.cpp)

```{r eval=F}
#https://github.com/barkasn/hogsvdR/blob/master/src/hogsvd.cpp
#include <RcppArmadillo.h>
#include <vector>

#ifdef _OPENMP
#include <omp.h>
#endif

// [[Rcpp::plugins(openmp)]]

using namespace Rcpp;

// [[Rcpp::depends(RcppArmadillo)]]
```


If directly use `sourceCpp('hogsvd.cpp')`, there would be a error message like "couldn't find openMP.

I did some serach, turned out this is a 'universal' problem for Apple and could be solved by some [system configuraitons](https://stackoverflow.com/questions/65251887/clang-7-error-linker-command-failed-with-exit-code-1-for-macos-big-sur/65334247#65334247).

Knowledge borrowed from stackoverflow:

1. Install xcode (if you have xcode, skip this step)

```{r eval=F}
# To delete an existing command line tools installation:
sudo rm -rf /Library/Developer/CommandLineTools
# To install the command line tools
sudo xcode-select --install
# check your xcode-select version
xcode-select --version
# xcode-select version 2354.
```

2. Install gcc via Homebrew (if you have gcc, skip this step)

```{r eval=F}
brew install gcc
# check your gcc version
gcc --version

Configured with: --prefix=/Library/Developer/CommandLineTools/usr --with-gxx-include-dir=/Library/Developer/CommandLineTools/SDKs/MacOSX10.14.sdk/usr/include/c++/4.2.1
Apple LLVM version 10.0.1 (clang-1001.0.46.4)
Target: x86_64-apple-darwin18.6.0
Thread model: posix
InstalledDir: /Library/Developer/CommandLineTools/usr/bin
```

3. link some headers into /usr/local/include

```{r eval=F}
sudo ln -s /Library/Developer/CommandLineTools/SDKs/MacOSX.sdk/usr/include/* /usr/local/include/
```

4. check your gfortran version and edit `~/.R/Makevars` file.

I follow [this answer](https://stackoverflow.com/questions/68263165/installing-r-on-osx-big-sur-edit-and-apple-m1-for-use-with-rcpp-and-openmp) to modify my `Makevars` file


```{r eval=F}
LOC = /usr/local/gfortran
CC=$(LOC)/bin/gcc -fopenmp
CXX=$(LOC)/bin/g++ -fopenmp
CXX11 = $(LOC)/bin/g++ -fopenmp

CFLAGS=-g -O3 -Wall -pedantic -std=gnu99 -mtune=native -pipe
CXXFLAGS=-g -O3 -Wall -pedantic -std=c++11 -mtune=native -pipe
LDFLAGS=-L$(LOC)/lib -Wl,-rpath,$(LOC)/lib
# find your SDK path via '$xcrun -show-sdk-path' 
# https://firas.io/post/data.table_openmp/
CPPFLAGS=-I$(LOC)/include -I/Library/Developer/CommandLineTools/SDKs/MacOSX10.14.sdk/usr/include
#CPPFLAGS=-I$(LOC)/include -I/Library/Developer/CommandLineTools/SDKs/MacOSX.sdk/usr/include

# (check that the version of gfortran - in this case 10.2.0 - matches the version specified in FLIBS)
FLIBS=-L/usr/local/gfortran/lib/gcc/x86_64-apple-darwin18/8.2.0 -L/usr/local/gfortran/lib -lgfortran -lquadmath -lm 
#FLIBS=-L/usr/local/gfortran/lib/gcc/x86_64-apple-darwin19/10.2.0 -L/usr/local/gfortran/lib -lgfortran -lquadmath -lm
CXX1X=/usr/local/gfortran/bin/g++
CXX98=/usr/local/gfortran/bin/g++
CXX11=/usr/local/gfortran/bin/g++
CXX14=/usr/local/gfortran/bin/g++
CXX17=/usr/local/gfortran/bin/g++
```

Then, open you Rstudio, check if openMP is now enabled.

# Validation: openMP installation

## Using data.table with OpenMP support

My previous installed `data.table` R package was not compiled in presnce of `OpenMP`. Thus, whenever I `library(data.table)`, message like 

> This installation of data.table has not detected OpenMP support.
> It should still work but in single-threaded mode.
would show up.


Now I'd [re-install `data.table` to enable OpenMP capacity](https://firas.io/post/data.table_openmp/).

```{r eval=F}
remove.packages("data.table")
install.packages("data.table", type = "source",
    repos = "https://Rdatatable.gitlab.io/data.table")
```

To check if `data.table` was successfully compiled with OpenMP support, load the library:

```{r eval=F}
library(data.table)
# data.table 1.14.0 using 4 threads (see ?getDTthreads).  Latest news: r-datatable.com
```

## Rcpp sleep test
An alternative to test OpenMP is to [use the `sleep` function](https://mfasiolo.github.io/sc2-2019/rcpp_advanced_iii/1_openmp/).

**serial sleep**

```{r}
library(Rcpp)
sourceCpp(code = '
#include <unistd.h>
#include <Rcpp.h>

// [[Rcpp::export(wait_a_second)]]
bool wait_a_second(int sec)
{
 for(size_t ii = 0; ii < sec; ii++)
 { 
  sleep(1);
 }
 return 1;
}
')

system.time( wait_a_second(2) )[3]
```

**parallel sleep**

```{r}
sourceCpp(code = '
#include <unistd.h>
#include <Rcpp.h>

// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(wait_a_second_omp)]]
bool wait_a_second_omp(int sec, int ncores)
{

 #if defined(_OPENMP)
  #pragma omp parallel num_threads(ncores)
  #pragma omp for
 #endif
 for(size_t ii = 0; ii < sec; ii++)
 { 
  sleep(1);
 }
 
 return 1;

 }
')
#para1: sleep seconds, para2: n.core used
system.time(wait_a_second_omp(4, 1))
system.time(wait_a_second_omp(4, 4))
system.time(wait_a_second_omp(16, 16))
```

# A detour: benchmark code performance

It's a routine task to benchmark code running time on samll dataset before applying it to big project.

Here is a simple example with R base function `system.time` to if [`mclapply` is doing parallel computation as we expect](https://cosx.org/2016/09/r-and-parallel-computing/
).

```{r}
solve.quad.eq <- function(a, b, c) {
  if(abs(a) < 1e-8 && abs(b) < 1e-8) return(c(NA, NA) )
  if(abs(a) < 1e-8 && abs(b) > 1e-8) return(c(-c/b, NA))
  if(b*b - 4*a*c < 0) return(c(NA,NA))
  x.delta <- sqrt(b*b - 4*a*c)
  x1 <- (-b + x.delta)/(2*a)
  x2 <- (-b - x.delta)/(2*a)
  return(c(x1, x2))
}

library(parallel)

# Generate data 
len <- 1e6
a <- runif(len, -10, 10)
a[sample(len, 100,replace=TRUE)] <- 0
b <- runif(len, -10, 10)
c <- runif(len, -10, 10)
# serial code
system.time(
  res1.s <- lapply(1:len, FUN = function(x) { solve.quad.eq(a[x], b[x], c[x])})
)

# multicores on Linux or Mac
system.time(
  res1.p <- mclapply(1:len, FUN = function(x) { solve.quad.eq(a[x], b[x], c[x])}, mc.cores = 2)
)
```



# Rcpp application: large matrix multiplication
I saw [this example](https://stackoverflow.com/questions/35923787/fast-large-matrix-multiplication-in-r) on stackoverflow and copied it here showing the power of Rcpp.


```{r}
library(Rcpp)
sourceCpp(code ='
// [[Rcpp::depends(RcppArmadillo, RcppEigen)]]

#include <RcppArmadillo.h>
#include <RcppEigen.h>

// [[Rcpp::export]]
SEXP armaMatMult(arma::mat A, arma::mat B){
    arma::mat C = A * B;

    return Rcpp::wrap(C);
}

// [[Rcpp::export]]
SEXP eigenMatMult(Eigen::MatrixXd A, Eigen::MatrixXd B){
    Eigen::MatrixXd C = A * B;

    return Rcpp::wrap(C);
}

// [[Rcpp::export]]
SEXP eigenMapMatMult(const Eigen::Map<Eigen::MatrixXd> A, Eigen::Map<Eigen::MatrixXd> B){
    Eigen::MatrixXd C = A * B;

    return Rcpp::wrap(C);
}
')


A <- matrix(rnorm(10000), 100, 100)
B <- matrix(rnorm(10000), 100, 100)

library(microbenchmark)
microbenchmark(A%*%B, armaMatMult(A, B), eigenMatMult(A, B), eigenMapMatMult(A, B))

```

You can see the different running speed of differemt implementations of matrix multiplication, `A%*%B` is the slowest and `eigenMapMatMult(A, B)` is the fastest.

**References**

1) https://mac.r-project.org/openmp/
2) https://github.com/barkasn/hogsvdR/blob/master/src/hogsvd.cpp
3) https://stackoverflow.com/questions/65251887/clang-7-error-linker-command-failed-with-exit-code-1-for-macos-big-sur/65334247#65334247
4) https://stackoverflow.com/questions/68263165/installing-r-on-osx-big-sur-edit-and-apple-m1-for-use-with-rcpp-and-openmp
5) https://firas.io/post/data.table_openmp/
6) https://mfasiolo.github.io/sc2-2019/rcpp_advanced_iii/1_openmp/
7) https://cosx.org/2016/09/r-and-parallel-computing/
8) https://stackoverflow.com/questions/35923787/fast-large-matrix-multiplication-in-r


```{r}
sessionInfo()
```

