---
title: "03.scRNA-seq: fit P, NB, ZIP, ZINB models in a Bayesian framework with simulated data"
description: |
  Simulate data following the four statistical models and estimate parameters in a Bayesian framework with R package rstanarm, brms, and cmdstanr.
author: Ming
date: 2021-11-23
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this post, I'd fit the four models in a Bayesian framework with simulated data.

## set up formulas
```{r}
# simulate data and fit model using rstan
library(rstanarm)
#https://mc-stan.org/rstanarm/reference/stan_glm.html

f0='y ~ 1' #intercept only model
f1='y ~ 1+group' #with one covariate
nCores=2;
n=1000;

seed=2049;
group <- sample(c(0,1), size = n, replace = TRUE) #covariate
```

## Poisson model
```{r}

# simulate data following Poisson model and fit model
seed=2049;
simu.dat=data.frame(y=rpois(n,lambda=2),group=group)

m0 <-   stan_glm(as.formula(f0),
                             family = poisson,#default link = "log" 
                             data = simu.dat,
                             cores = nCores,
                             seed = seed,
                             refresh = 0)
m0$coefficients
exp(m0$coefficients)

m1 <-   stan_glm(as.formula(f1),
                 family = poisson,#default link = "log" 
                 data = simu.dat,
                 cores = nCores,
                 seed = seed,
                 refresh = 0)
m1$coefficients
exp(m1$coefficients) 

# use loo to compare two models
#http://mc-stan.org/loo/reference/loo_compare
#https://github.com/churchill-lab/scRATE/blob/master/R/compare_count_models.R
#https://mc-stan.org/loo/reference/loo-glossary.html
#The ELPD is the theoretical expected log pointwise predictive density for a new dataset (Eq 1 in VGG2017), which can be estimated, e.g., using cross-validation. elpd_loo is the Bayesian LOO estimate of the expected log pointwise predictive density (Eq 4 in VGG2017) and is a sum of N individual pointwise log predictive densities. 

library(loo)
loo_list=list()
loo_list[['m0']]=loo(m0)
loo_list[['m1']]=loo(m1)
elpd_loo <- do.call(loo_compare, list(loo_list))
elpd_loo
#the best model is in the 1st row
```

## Negative Binomial model 
```{r}

# simulate data following Negative Binomial model and fit model
#http://mc-stan.org/rstanarm/reference/neg_binomial_2.html
#https://mc-stan.org/rstanarm/reference/stan_glm.html
#For negative binomial models prior_aux controls "reciprocal_dispersion", 
#which is similar to the "size" parameter of rnbinom: 
#smaller values of "reciprocal_dispersion" correspond to greater dispersion. 

## 1) test model
seed=2049
simu.dat=data.frame(y=MASS::rnegbin(1000,mu=2,theta=0.5),group=group)
seed=2049
simu.dat=data.frame(y=rnbinom(1000,mu=2,size=0.5),group=group)

m0=stan_glm(as.formula(f0),
            family = neg_binomial_2, #default, link='log'
            data = simu.dat,
            cores = nCores,
            seed = seed,
            refresh = 0)

m0
m0$coefficient
exp(m0$coefficients) #default, link='log'

## 2) model comparison
simu.dat=data.frame(y=MASS::rnegbin(1000,mu=exp(2+1.5*(group==1)),
                                    theta=0.5),group=group)

m0=stan_glm(as.formula(f0),
            family = neg_binomial_2, #default, link='log'
            data = simu.dat,
            cores = nCores,
            seed = seed,
            refresh = 0)

m0
m0$coefficient
#exp(fit.out$coefficients) #default, link='log'

m1=stan_glm(as.formula(f1),
            family = neg_binomial_2, #default, link='log'
            data = simu.dat,
            cores = nCores,
            seed = seed,
            refresh = 0)
m1
m1$coefficients

loo_list[['m0']]=loo(m0)
loo_list[['m1']]=loo(m1)
loo_compare(loo_list)
```

## Zero-inflated Poisson model

A Zero-inflated Possion model has two component: inlfated 0 and a Possion distribution.

Thus, the probability of observing a 0 equals:

$$ P(Y=0) = \pi + (1-\pi) *Poisson(0|lambda) $$


```{r}

# use brms package to fit zero-inflated models
library(brms);
adapt_delta=0.8;

options(brms.backend = "cmdstanr")

# simulate data following zero-inflated Possion model and fit model
#https://stats.stackexchange.com/questions/189005/simulate-from-a-zero-inflated-poisson-distribution/189052
pi=0.2; lambda=2
y=ifelse(rbinom(n, size = 1, prob = 1-pi) == 0, 0, rpois(n, lambda = lambda))
simu.dat=data.frame(y=y)

#https://stat.ethz.ch/R-manual/R-devel/library/stats/html/NegBinomial.html
myprior <- get_prior(bf(as.formula(f0), zi ~ 1),
                       family = zero_inflated_poisson(),
                       data = simu.dat)
myprior_values <- eval(parse(text=gsub("student_t", "c", myprior$prior[1])))
m0<- brm(bf(as.formula(f0), zi ~ 1),
                        family = zero_inflated_poisson(),
                        data = simu.dat,
                        prior = myprior,
                        control = list(adapt_delta = adapt_delta),
                        cores = nCores,
                        seed = seed,
                        refresh = 500)
m0
x=brms::posterior_summary(m0)
x
# possion, lambda, link function: mu=log
exp(x[1,1]) #lambda
# zero inflation, link function: zi = logit 
exp(x[4,1])/(1+exp(x[4,1])) #p
```

## Zero-inflated Negative Binomial model

A Zero-inflated Negative Binomial model has two component: inlfated 0 and a Negative Binomial distribution.

Thus, the probability of observing a 0 equals:

$$ P(Y=0) = \pi + (1-\pi) *NB(0|mu,size) $$

```{r}

# simulate data following zero-inflated Negative Binomial model and fit model
set.seed(6)
pi=0.3
z <- rbinom(n = n, size = 1, prob = 1-pi) 
# mean(z == 0)
y_sim <- ifelse(z == 0, 0, 
                rnbinom(n = n, 
                        mu = exp(5), 
                        size = 3))
simu.dat=data.frame(y=y_sim,group=group)

message('Fitting data with Zero-Inflated Negative Binomial model...')
myprior <- get_prior(bf(as.formula(f0), zi ~ 1),
                       family = zero_inflated_negbinomial(),
                       data = simu.dat)
myprior_values <- eval(parse(text=gsub("student_t", "c", myprior$prior[1])))
m0 <- brm(bf(as.formula(f0), zi ~ 1),
                         family = zero_inflated_negbinomial(),
                         data = simu.dat,
                         control = list(adapt_delta = adapt_delta),
                         prior = myprior,
                         cores = nCores,
                         seed = seed,
                         refresh = 500)
m0
#Family: zero_inflated_negbinomial 
#Links: mu = log; shape = identity; zi = logit 
x=posterior_summary(m0)
x
# possion, lambda, link function: mu=log
x[1,1] #lambda
x[3,1] #shape or size
# zero inflation, link function: zi = logit 
exp(x[5,1])/(1+exp(x[5,1])) #prob

## add covarite
set.seed(6)
pi=0.3
z <- rbinom(n = n, size = 1, prob = 1-pi) 
# mean(z == 0)
y_sim <- ifelse(z == 0, 0, 
                rnbinom(n = n, 
                        mu = exp(5+1.5*(group==1)), 
                        size = 3))
simu.dat=data.frame(y=y_sim,group=group)

message('Fitting data with Zero-Inflated Negative Binomial model...')
myprior <- get_prior(bf(as.formula('y~1'), zi ~ 1),
                       family = zero_inflated_negbinomial(),
                       data = simu.dat)
myprior_values <- eval(parse(text=gsub("student_t", "c", myprior$prior[1])))
m0<-brm(bf(as.formula('y~1'), zi ~ 1),
        family = zero_inflated_negbinomial(),
        data = simu.dat,
        control = list(adapt_delta = adapt_delta),
        prior = myprior,
        cores = nCores,
        seed = seed,
        refresh = 500)
x=brms::posterior_summary(m0)
x

myprior <- get_prior(bf(as.formula('y~1+group'), zi ~ 1),
                     family = zero_inflated_negbinomial(),
                     data = simu.dat)
m1 <- brm(bf(as.formula('y~1+group'), zi ~ 1),
                         family = zero_inflated_negbinomial(),
                         data = simu.dat,
                         control = list(adapt_delta = adapt_delta),
                         prior = myprior,
                         cores = nCores,
                         seed = seed,
                         refresh = 500)


x=brms::posterior_summary(m1)
x
#Family: zero_inflated_negbinomial 
#Links: mu = log; shape = identity; zi = logit 
# possion, lambda, link function: mu=log
x[1,1] #lambda
x[4,1] #shape or size
x[3,1] #covariate beta
# zero inflation, link function: zi = logit 
exp(x[6,1])/(1+exp(x[6,1])) #prob

loo_list=list()
loo_list[['m0']]=loo(m0)
loo_list[['m1']]=loo(m1)
loo_compare(loo_list)
```

```{r}
sessionInfo()
```






